---
title: "Project 1: Olmypic History Trend Analytics"
author: "Shruti"
output: pdf_document
always_allow_html: true

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, message=FALSE}
#Importing Required packages
library(tidyverse)
library(reshape2)
library(dplyr)
library(knitr)
library(gridExtra)
library(ggplot2)
library(data.table)

```



## Data Description

 120 years of Olympic history. 
 There are two files in our data set:

*athlete_events.csv* contains 271116 rows and 15 columns. 

- ID : Unique number for each athlete
- Name : Athlete's name
- Sex : M or F
- Age : Integer
- Height : In centimeters
- Weight : In kilograms
- Team : Team name
- NOC : National Olympic Committee 3-letter code
- Games : Year and season
- Year : Integer
- Season : Summer or Winter
- City : Host city
- Sport : Sport
- Event : Event
- Medal : Gold, Silver, Bronze, or NA

*noc_regions.csv* contains 230 rows and 3 columns.

- NOC : National Olympic Committee 3 letter code
- Country name : matches with regions in map_data("world")
- Notes : Special notes if any

data set from Kaggle: https://www.kaggle.com/heesoo37/120-years-of-olympic-history-athletes-and-results?select=athlete_events.csv


```{r, echo=FALSE} 
# Data of the athelets and countries
atheletes_df <- read.csv('athlete_events.csv', header = TRUE, sep = ',')
```

```{r, echo=FALSE}
# Importing Data of the regions tied with the NOC code
regions_df <- read.csv('noc_regions.csv', header= TRUE, sep =',')
```


# Data Wrangling

summary of the athletes data

```{r, echo=FALSE}
summary(atheletes_df)
```

summary of region data

```{r, echo=FALSE}
summary(regions_df)
```


```{r, echo=FALSE}
kable(
  atheletes_df %>%
    summarise(total_records=n()),
  caption = "Total Records for Athletes Dataframe"
  
)
```



```{r, echo=FALSE}
kable(
  regions_df %>%
    summarise(total_records=n()),
  caption = "Total Records in Regions Dataframe"
  
)
```

Looking for NA's 


```{r, echo=FALSE}
# Store the cols with missing values 
list_na <- colnames(atheletes_df)[apply(atheletes_df, 2, anyNA)]
list_na
```

We have NA's for numerical data: Age, Height & Weight and for categorical data: Medal.


```{r, echo=FALSE}
kable(
  atheletes_df %>%
    group_by(Medal) %>%
    summarise(total_records=n())
    ,caption="Records by Medal Count"
  
)
```
There are 231333 NA's for Medals 

```{r, echo=FALSE, message=FALSE}
#looking for NA's in regions_df
kable(
  regions_df %>%
    filter(is.na(region)) %>%
    group_by(NOC,region,notes) %>%
  summarise(Total_records=n()),
  caption="Records grouped by categories"
)
```

There are 3 records with NA in region_df



### A. Handling Missing Data

#### B. Replacing NA's in Medals

```{r, echo=FALSE}
atheletes_df$Medal <- atheletes_df$Medal %>% 
  replace_na("None") 
```

Calculating  Median for the missing values for Age, Height and Weight

```{r, echo=FALSE}

list_na <- list_na[ list_na !="Medal"]

# Calculate median for the missing values
missing_median <- apply(atheletes_df[,colnames(atheletes_df) %in% list_na],
      2, # 2 is for Columns
      median,
      na.rm =  TRUE)
missing_median
```

```{r, echo=FALSE}

atheletes_df <- atheletes_df %>%
   mutate(
     Age  = ifelse(is.na(Age), missing_median[1], Age),
     Height = ifelse(is.na(Height), missing_median[2], Height),
     Weight = ifelse(is.na(Weight), missing_median[3], Weight)
   )
```

#### C. Handling Missing data in Regions

 Na's values with region/notes for respective NOC's for the region data.
```{r, echo=FALSE}

regions_df$region <- ifelse(is.na(regions_df$region), regions_df$notes, regions_df$region)
```


```{r, echo=FALSE}
kable(
  regions_df %>%
    filter(is.na(region)) %>%
    group_by(region) %>%
    summarise(total_records=n())
  ,caption = "Number of NA's in Region after fix"
)

```



 join regions_df and atheletes_df based on the NOC code.


```{r, echo=FALSE, results='hide'}
athletes <- left_join(atheletes_df, regions_df, by="NOC")


colnames(athletes)[which(names(athletes) == "region")] <- "Region" 

athletes <- athletes[,-17]
head(athletes,5)

```


### A. Check for any missing values

```{r, echo=FALSE, message=FALSE}

 # Counting the number of NA's for all the columns
 colnames(athletes)[apply(athletes, 2, anyNA)]

kable(
  athletes %>%
    select(NOC,Region)  %>%
    filter(is.na(Region)) %>%
    group_by(NOC,Region) %>%
    summarise(total_records=n())
    ,caption="Null Records check by Medal Count"
  
)

```
For NOC SGP, there are no records in our regions_df but is present in atheltes_df, as a result we are getting NA values after the join. We will add Singapore Region to the NOC in the joined data. Applying fix:

```{r, echo=FALSE}

athletes$Region <- ifelse((is.na(athletes$Region) & athletes$NOC=='SGP'), "Singapore", athletes$Region)


kable(
  athletes %>%
    select(Region)  %>%
    filter(is.na(Region)) %>%
    group_by(Region) %>%
    summarise(total_records=n())
    ,caption="Checking for NA records in Region after change"
  
)

```

 Check for Duplicates

```{r, echo=FALSE}
sum(duplicated(athletes))
```
There **1385** duplicate records on the whole data set. Taking unique values to remove duplicates.


```{r, echo=FALSE}
# Removing the duplicates
athletes <- unique(athletes)

```


### C. Checking boundary cases

```{r, echo=FALSE}
kable (
athletes %>%
summarise(max_age=max(Age), min_age=min(Age), Average_Age=mean(Age)),
caption="Age boundary cases"

)

kable (
athletes %>%
summarise(max_height=max(Height), min_height=min(Height), Average_height=mean(Height)),
caption="Height boundary cases"

)

kable (
athletes %>%
summarise(max_weight=max(Weight), min_weigt=min(Weight), Average_weight=mean(Weight)),
caption="Weight boundary cases"

)
```
All our boundary cases looks reasonable and accurate after our wrangling.



#  Analytics

### Top regions with the highest number of medals 


```{r}

library(ggplot2)

# Create a custom theme for the plots
custom_theme <- theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 12),
    legend.position = "top",
    legend.title = element_blank(),
    legend.text = element_text(size = 12),
    panel.grid.major = element_line(colour = "lightgray", size = 0.2),
    panel.grid.minor = element_blank()
  )

# Filter and summarize the data for 1896-1956
top_regions_1896_1956 <- athletes %>%
  filter(Medal != 'None' & Year <= 1956) %>%
  group_by(Region) %>%
  summarize(total_medals = n()) %>%
  arrange(desc(total_medals)) %>%
  mutate(Region = factor(Region, levels = Region)) %>%
  slice(1:10)

# Create a bar chart for 1896-1956 with custom styling and color
p1 <- ggplot(top_regions_1896_1956, aes(x = Region, y = total_medals)) +
  geom_col(fill = "steelblue") +
  labs(
    y = "Number of Medals",
    x = "",
    title = "Top 10 Regions with Highest Medals (1896-1956)"
  ) +
  custom_theme +
  geom_vline(xintercept = 0, color = "gray") +
  geom_hline(yintercept = 0, color = "gray")

# Filter and summarize the data for 1957-2016
top_regions_1957_2016 <- athletes %>%
  filter(Medal != 'None' & Year > 1956) %>%
  group_by(Region) %>%
  summarize(total_medals = n()) %>%
  arrange(desc(total_medals)) %>%
  mutate(Region = factor(Region, levels = Region)) %>%
  slice(1:10)

# Create a bar chart for 1957-2016 with custom styling and different bar color
p2 <- ggplot(top_regions_1957_2016, aes(x = Region, y = total_medals)) +
  geom_col(fill = "lightcoral") +
  labs(
    y = "Number of Medals",
    x = "",
    title = "Top 10 Regions with Highest Medals (1957-2016)"
  ) +
  custom_theme +
  geom_vline(xintercept = 0, color = "gray") +
  geom_hline(yintercept = 0, color = "gray")

# Display the two graphs separately
grid.arrange(p1, p2, ncol = 1)


```

###  2. Medals won by Males/Females 

```{r}
library(ggplot2)

# Create a custom theme for the plot
custom_theme <- theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 12),
    legend.position = "top",
    legend.title = element_blank(),
    legend.text = element_text(size = 12),
    panel.grid.major = element_line(colour = "lightgray", size = 0.2),
    panel.grid.minor = element_blank()
  )

# Filter and summarize the data
medals_by_year_sex <- athletes %>%
  filter(Medal != 'None') %>%
  group_by(Year, Sex) %>%
  summarize(Total_Medals = n())

# Create an area chart for cumulative medals won by Male/Female over time
medals_over_time_area_plot <- ggplot(medals_by_year_sex, aes(x = Year, y = cumsum(Total_Medals), fill = Sex)) +
  geom_area() +
  scale_fill_manual(values = c("darkblue", "red")) +
  labs(
    title = "Cumulative Medals won by Male/Female over Time",
    x = "Year",
    y = "Cumulative Medals"
  ) +
  custom_theme

medals_over_time_area_plot

```

###  3.  most participated sport in Olympics every year

```{r, echo=FALSE, message=FALSE}

q3<-
athletes %>% 
  group_by(Year, Sport) %>%
  summarize(Participation = n()) %>%
  arrange(Year,desc(Participation))
  
  q3 <- q3[!duplicated(q3$Year),] # Taking only the first record
  
  kable(
  q3,
  caption="Most participated Sport in Olympic Games every year"
  )

```

### 4.  country win a medal for the first time for a particular sport by year

```{r, echo=FALSE, message=FALSE}
Ans1 <-
athletes %>%
filter(Sport=="Gymnastics",Medal!="None") %>%
select(Region,Year) %>%
group_by(Region,Year) %>%
summarise(Year=min(Year))

Ans1 <- Ans1[!duplicated(Ans1$Region),]

```

```{r}
# Load the required libraries
library(kableExtra)
library(dplyr)

# Create a styled table with two columns
styled_table <- Ans1 %>%
  kable("html", escape = FALSE, align = "c") %>%
  kable_styling("striped", full_width = FALSE) %>%
  row_spec(0, bold = TRUE) %>%
  add_header_above(c("Country" = 1, "Year" = 1)) %>%
  column_spec(1, bold = TRUE) %>%
  collapse_rows(columns = 1, valign = "middle") %>%
  column_spec(2, width = "2em")

styled_table


```


### 5.  per sport per country for the gap between medals for Gymnastics


```{r}
# Load the required library
library(ggplot2)

# Define a custom bar color
custom_bar_color <- "pink"

athletes %>%
  filter(Sport == "Gymnastics", Medal != "None") %>%
  select(Region, Year) %>%
  group_by(Region) %>%
  summarise(Number_of_Years = max(Year) - min(Year)) %>%
  arrange(desc(Number_of_Years)) %>%
  mutate(Region = factor(Region, levels = Region)) %>%
  slice(1:20) %>%
  ggplot(aes(y = Region, x = Number_of_Years, label = Number_of_Years)) +
  geom_col(fill = custom_bar_color) +
  theme_minimal() +
  geom_text(hjust = -0.2, size = 3, color = "black") +
  labs(x = "Number of Years", y = "Region") +
  xlab("Number of Years") +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10)
  ) +
  ggtitle("Gap between medals for countries in Gymnastics")


```
### 6. medals won by countries in Summer & Winter.
```{r fig.width=10, echo=FALSE}
Ans3.1 <-
athletes %>%
filter(Medal!='None',Season=="Summer") %>%
group_by(Region) %>%
summarize(number_of_medals=n()) %>%
arrange(desc(number_of_medals)) %>%
mutate(Region=factor(Region, levels=Region)) %>%
slice(1:20) %>%
ggplot( aes(y=Region, x=number_of_medals))+
geom_col(fill="yellow") +
theme_minimal()+
#labs(y="Region")+
xlab("Medals")+
theme(plot.title = element_text(hjust = 0.5)) +
ggtitle("Number of medals won by countries in Summer")+
geom_vline(xintercept = 0)+
geom_hline(yintercept = 0)

Ans3.2 <-
athletes %>%
filter(Medal!='None',Season=="Winter") %>%
group_by(Region) %>%
summarize(number_of_medals=n()) %>%
arrange(desc(number_of_medals)) %>%
mutate(Region=factor(Region, levels=Region)) %>%
slice(1:20) %>%
ggplot( aes(y=Region, x=number_of_medals))+
geom_col(fill="darkblue") +
theme_minimal()+
#labs(y="Region")+
xlab("Medals")+
theme(plot.title = element_text(hjust = 0.5)) +
ggtitle("Number of medals won by countries in Winter")+
geom_vline(xintercept = 0)+
geom_hline(yintercept = 0)

grid.arrange(Ans3.1,Ans3.2,ncol=1)

```

### 7. Host cities with the highest participation.

```{r}
library(ggplot2)
library(dplyr)

# Filter and summarize data for the top host cities
Top10 <- athletes %>%
  select(Year, City) %>%
  group_by(Year, City) %>%
  summarize(number = n()) %>%
  arrange(desc(number)) %>%
  head(10)

# Calculate the maximum number of participants for setting X-axis limits
max_participants <- ceiling(max(Top10$number) / 2500) * 2500

# Create a horizontal bar chart with labels on the bars
ggplot(Top10, aes(x = number, y = reorder(City, -number))) +
  geom_bar(stat = "identity", fill = "brown") +
  geom_text(aes(label = paste("Year:", Year, ", Participants:", number),
                x = 0, hjust = 0),
            vjust = 0.5, size = 4, color = "white") +
  theme_minimal() +
  labs(x = "Number of Participants", y = "City") +
  scale_x_continuous(limits = c(0, max_participants), breaks = seq(0, max_participants, by = 2500)) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    axis.text.x = element_text(angle = 0, hjust = 0.5) # Rotate x-axis labels
  ) +
  ggtitle("Top 10 Host Cities with Highest Participation")



```
### 8.  athletes with the highest participation 

```{r}
library(ggplot2)
library(dplyr)

# Filter and summarize data for the top 10 most participated athletes
Top10a <- athletes %>%
  select(Name) %>%
  group_by(Name) %>%
  summarize(number = n())  %>%
  arrange(desc(number)) %>%
  slice(1:10)

# Create a horizontal bar chart with labels on the bars
ggplot(Top10a, aes(x = reorder(Name, number), y = number)) +
  geom_col(fill = "gold") +
  geom_text(aes(label = number), vjust = -0.2, size = 5, color = "black") +
  scale_y_continuous(limits = c(0, 80)) +  # Limit the Y-axis scale to 0 to 80
  theme_minimal() +
  labs(x = "Athletes", y = "Number of Olympics") +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1) # Rotate x-axis labels
  ) +
  ggtitle("Top 10 Most Participated Athletes in the Olympics") +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = 0)


```


### 9. Athletes with the most number of medals in each sport


```{r}
library(dplyr)

# Filter and summarize data to find the athlete with the most medals in each sport
df <- athletes %>%
  filter(Medal != 'None') %>%
  select(Name, Sport) %>%
  group_by(Name, Sport) %>%
  summarize(number = n()) %>%
  arrange(desc(number))

# Find the athlete with the most medals in each sport
df_top_athletes <- df %>% group_by(Sport) %>% top_n(1, number)

# Create a data frame for the table
table_data <- df_top_athletes %>%
  select(Sport, Name, number) %>%
  rename(Sport = "Sport", Athlete = "Name", Medals = "number")

# Print the table
kable(
  table_data,
  caption = "Athletes with the Most Number of Medals in Each Sport"
)


```

### 10. frequency of Olympics hosted by cities


```{r}
# Filter and summarize data for the frequency of Olympics hosted by cities
host_cities_data <- athletes %>%
  group_by(City) %>%
  summarize(Olympics_Hosted = n_distinct(Year))

# Create a bar chart to visualize the frequency of Olympics hosted by different cities
host_cities_chart <- ggplot(host_cities_data, aes(x = City, y = Olympics_Hosted)) +
  geom_col() +
  labs(
    x = "City",
    y = "Number of Olympics Hosted",
    title = "Frequency of Olympics Hosted by Different Cities"
  ) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

host_cities_chart




```
### 11. countries with most medals
```{r}
# Filter and summarize data for the top 5 countries with the most medals
top5_countries_most_medals <- athletes %>%
  filter(Medal != 'None') %>%
  group_by(Region) %>%
  summarize(Total_Medals = n()) %>%
  arrange(desc(Total_Medals)) %>%
  slice(1:5)

# Create a pie chart to show the distribution of medals among the top 5 countries
library(ggplot2)
library(dplyr)

top5_countries_most_medals$Region <- factor(top5_countries_most_medals$Region, levels = top5_countries_most_medals$Region)

# Calculate the percentage of medals for each country
total_medals <- sum(top5_countries_most_medals$Total_Medals)
top5_countries_most_medals <- top5_countries_most_medals %>%
  mutate(Percentage = (Total_Medals / total_medals) * 100)

# Create a pie chart with counts to show the distribution of medals among the top 5 countries with the most medals
ggplot(top5_countries_most_medals, aes(x = "", y = Percentage, fill = Region)) +
  geom_bar(stat = "identity", width = 1) +
  geom_text(aes(label = Total_Medals), position = position_stack(vjust = 0.5)) +
  coord_polar(theta = "y") +
  labs(title = "Distribution of Medals Among Top 5 Countries",
       fill = "Country") +
  scale_fill_manual(values = c("gold", "green", "pink", "orange", "purple")) +
  theme_void()

```
###12. 5 countries with the least medals
```{r}
# Filter and summarize data for the top 5 countries with the least medals
top5_countries_least_medals <- athletes %>%
  filter(Medal != 'None') %>%
  group_by(Region) %>%
  summarize(Total_Medals = n()) %>%
  arrange(Total_Medals) %>%
  slice(1:5)

# Create a pie chart to show the distribution of medals among the top 5 countries
library(ggplot2)
library(dplyr)

top5_countries_least_medals$Region <- factor(top5_countries_least_medals$Region, levels = top5_countries_least_medals$Region)

# Calculate the percentage of medals for each country
total_medals_least <- sum(top5_countries_least_medals$Total_Medals)
top5_countries_least_medals <- top5_countries_least_medals %>%
  mutate(Percentage = (Total_Medals / total_medals_least) * 100)

# Create a pie chart with counts to show the distribution of medals among the top 5 countries with the least medals
ggplot(top5_countries_least_medals, aes(x = "", y = Percentage, fill = Region)) +
  geom_bar(stat = "identity", width = 1) +
  geom_text(aes(label = Total_Medals), position = position_stack(vjust = 0.5)) +
  coord_polar(theta = "y") +
  labs(title = "Distribution of Medals Among Top 5 Countries with Least Medals",
       fill = "Country") +
  scale_fill_manual(values = c("gold", "green", "brown", "gray", "blue")) +
  theme_void()

```
### 13. most famous sports

```{r}

# Filter and summarize data to find the top 10 most famous sports at Olympics
top10_famous_sports <- athletes %>%
  filter(Medal != 'None') %>%
  select(Sport) %>%
  group_by(Sport) %>%
  summarize(Total_Medals = n()) %>%
  arrange(desc(Total_Medals)) %>%
  slice(1:10)

# Create a bar chart to show the top 10 most famous sports at Olympics
library(ggplot2)

top10_famous_sports$Sport <- factor(top10_famous_sports$Sport, levels = top10_famous_sports$Sport)

ggplot(top10_famous_sports, aes(x = Sport, y = Total_Medals)) +
  geom_col(fill = "skyblue") +
  labs(title = "Top 10 Most Famous Sports at Olympics",
       x = "Sport",
       y = "Total Medals") 
  theme_minimal

```

### 14.least played sports
```{r}
library(dplyr)
library(ggplot2)

# Filter and summarize data to find the least played sports at the Olympics
least_played_sports <- athletes %>%
  filter(Medal != 'None') %>%
  select(Sport) %>%
  group_by(Sport) %>%
  summarize(Count = n()) %>%
  arrange(Count) %>%
  slice(1:10)

# Create a bar chart for the least played sports
ggplot(least_played_sports, aes(x = reorder(Sport, Count), y = Count)) +
  geom_col(fill = "lightcoral") +
  geom_text(aes(label = Count), vjust = -0.2, size = 5, color = "black") +
  theme_minimal() +
  labs(x = "Sport", y = "Count") +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  ggtitle("Top 10 Least Played Sports at the Olympics")

```
### 15. Gender Distribution 
```{r}

library(ggplot2)

# Assuming you have a data frame `atheletes_df` with columns "Year" and "Sex"
ggplot(atheletes_df, aes(x = Year, fill = Sex)) +
  geom_bar() +
  labs(title = "Gender Distribution Over the Years",
       x = "Olympic Year",
       y = "Number of Athletes")


```
### 16. Box Plot Comparisons by Gender
```{r}

library(ggplot2)

# Assuming you have a data frame `df` with columns "Characteristic" (e.g., "Age", "Height", "Weight") and "Gender"
ggplot(atheletes_df, aes(x = Sex, y = Age, fill = Sex)) +
  geom_boxplot() +
  labs(title = "Box Plot Comparisons by Gender",
       x = "Gender",
       y = "Characteristic")

ggplot(atheletes_df, aes(x = Sex, y = Weight, fill = Sex)) +
  geom_boxplot() +
  labs(title = "Box Plot Comparisons by Gender",
       x = "Gender",
       y = "Characteristic")

ggplot(atheletes_df, aes(x = Sex, y = Height, fill = Sex)) +
  geom_boxplot() +
  labs(title = "Box Plot Comparisons by Gender",
       x = "Gender",
       y = "Characteristic")



```
